---
title: "Bill tabulation"
date:  "`r format(Sys.Date(), '%B %d, %Y')`"
author: "Bijay Paudyal, Economist Intern and 
Shawn Stoddard, Senior Resource Economist"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

<style>
body {
  font-size: 14px;
}




```{r setup, include=FALSE}
### Setting up global knitr chunk to control code display options

knitr::opts_chunk$set(echo = FALSE,
                      eval=TRUE,
                      warning = FALSE,
                      results= 'asis',
                      error=FALSE)
```




```{r clear_work_space}
# Clearing previous work-space
rm(list=ls())
```





```{r load_library}
### Loading the library

library(tidyverse)
library(ggplot2)
library(knitr)
library(nls2)
```

</style>

```{r load_data}
#loading the data

df <- data.frame(readRDS("S:/2023/BPaudyal/Customer_Data/SingleFamilyUsage.rds"))

#changing the column names to lowercase
colnames(df)<-tolower(colnames(df))


```


The summary of original data is:
```{r summary_original_data}
#looking at the summary stat of our original data
summary_stats <- df %>%
                  summarize(
                            Minimum = min(tot_use),
                            Q1 = quantile(tot_use, 0.25),
                            Median = median(tot_use),
                            Q3 = quantile(tot_use, 0.75),
                            Maximum = max(tot_use)
                         )



kable(summary_stats,caption="Data summary of original data",
          align = 'c')

```


Reviewing the data to see the outliers using the box plot

```{r Checking outliers}
#reviewing the data to see the outliers using the boxplot

ggplot(df,aes(y=tot_use, x=rclass))+
  geom_boxplot()
```



```{r removing negative water use}
#creating the new data frame by removing the negative water use in the original data

df1<-df %>% 
    filter(tot_use>=0)


```



Summary after removing the negative water use :
```{r summary after neg removal}
summary_stats <- df1 %>%
                  summarize(
                            Minimum = min(tot_use),
                            Q1 = quantile(tot_use, 0.25),
                            Median = median(tot_use),
                            Q3 = quantile(tot_use, 0.75),
                            Maximum = max(tot_use)
                         )



kable(summary_stats,caption="Data summary after removing negative water use values",
          align = 'c')
```




Summary of the data after removing the negative water use and the top 1 percent outliers:

```{r removing top 1 percent}
#removing the top 1 percent of outliers (high water users)
df1<-df1 %>%
          filter(tot_use<=quantile(tot_use,0.99))




#summary_df1<-summary(df1$tot_use)


summary_stats <- df1 %>%
                  summarize(
                            Minimum = min(tot_use),
                            Q1 = quantile(tot_use, 0.25),
                            Median = median(tot_use),
                            Q3 = quantile(tot_use, 0.75),
                            Maximum = max(tot_use)
                         )



kable(summary_stats,caption="Data summary after removing top 1 percent outliers",
          align = 'c')
```



The boxplot of the clean data:

```{r box plot of clean data}
#the boxplot of the clean data after removing top and bottom outliers
df1 %>%
  ggplot(aes(x = rclass, y = tot_use)) +
  geom_boxplot(fill = "lightblue") +
  
  labs(title = "Total water usage Boxplot",
       x = "rclass",
       y = "Total use") +
 
    theme_minimal() 
  
```

The histogram of the data where x axis shows the water use and y axis shows the frequency
```{r histogram}
#creating the histogram of the clean data where x axis shows the water use and y axis shows the frequency
ggplot(df1, aes(x = tot_use)) +
  geom_histogram(fill = "lightgreen", binwidth=2) +
  labs(
    title = "Histogram of total water use",
    x = "Total Use"
  ) +
  facet_wrap(vars(rclass))+
  theme_minimal()

```




```{r tabulation sheet}
#Creating  a tabulation sheet
water_usage_values <- 0:max(df1$tot_use)
tabulation_sheet <- data.frame(water_usage = water_usage_values,
                               number_of_bills_ending_in_block = rep(0, length(water_usage_values)),
                               total_water_usage = rep(0, length(water_usage_values)))
```


##### applyng filter (Only RMWS)

```{r bill tabulation sheet}
#creating bill tabulating sheet
usage_counts <- df1 %>%
  #filter(rclass=="RMWD"|rclass=="RMWS") %>% 
  #filter(ssize=='0075') %>% 
  filter(rclass=="RMWS") %>% 
  group_by(tot_use) %>%
  summarize(total_bills = n())
```

```{r adding data}
#addding data to bill tabulation sheet
for (i in 1:nrow(usage_counts)) {
      
      
      use_value <- usage_counts$tot_use[i]
          
      count <- usage_counts$total_bills[i]
  
      tabulation_sheet <- tabulation_sheet %>%
                            mutate(number_of_bills_ending_in_block = ifelse(water_usage == use_value, 
                                                                  count, number_of_bills_ending_in_block))

}

```




```{r total water use for single customer}

#finding the total water usage for a single customer

tabulation_sheet<-tabulation_sheet %>% 
                    mutate(total_water_usage=water_usage*number_of_bills_ending_in_block)




total_usage<- tabulation_sheet %>%
              summarise(number_of_bills_ending_in_block = sum(number_of_bills_ending_in_block),
                        total_water_usage = sum(total_water_usage))


```


Development of cumulative billed usage


```{r cumulative billed usage}
#Development of cumulative billed usage

cumul_bill<-tabulation_sheet %>% 
                    select(-3)

```





```{r cumulative bills through block}
#get the cumulative sum from the largest block to get cumulative bills through block
cumul_bill<-cumul_bill %>% 
              mutate(cumul_bill_thru_block = rev(cumsum(rev(number_of_bills_ending_in_block))))
```


```{r total use of bills stopping in block}
#Column 4- total bills stopping in each block and corresponds to number of bills listed in column 2
cumul_bill<-cumul_bill %>% 
                      mutate(total_use_of_bill_stopping_in_block=water_usage*number_of_bills_ending_in_block)


```



```{r cumulative use of bills stopping in block}
#column5- cumulative use f bills stopping in block
cumul_bill <- cumul_bill %>%
 mutate(cumul_use_of_bills_stopping_in_block = lag(total_use_of_bill_stopping_in_block) + total_use_of_bill_stopping_in_block)

```



```{r total use to bills passing through block}
#column 6- total use to this block of all bills passing through block
cumul_bill<-cumul_bill %>% 
            mutate(total_use_to_thos_block_of_all_bills_passing_through_block = (water_usage) * lead(cumul_bill_thru_block))
```



```{r  cumulative billed usage final}
# column 7 : cumulative billed usage is the cumulative sum of cumulative billed usage through block
cumul_bill <- cumul_bill %>%
   mutate(cumulative_billed_usage = ifelse(row_number() == 1, 0, cumul_bill_thru_block)) %>% 
  mutate(cumulative_billed_usage=cumsum(cumulative_billed_usage))
```
  
```{r cumulative Billed Usage percent}

#column 8 : cumulative Billed Usage in terms of Percentage

cumul_bill <- cumul_bill %>%
  mutate(cumulative_billed_usage_percent = (cumulative_billed_usage / last(cumulative_billed_usage)) * 100) %>% 
  mutate(cumulative_billed_usage_percent=  round(cumulative_billed_usage_percent,2))

```



```{r billed usage table}
cumul_bill %>%
    kable(caption = "Development of cumulative billed usage", 
          col.names = c('Usage Block 1000 gal', 'Number of bills ending in Block', 'Cumulative bills through Block','Total use of Bills Stopping in Block, 1000 gal', 'Cumulative use of Bills Stopping in Block, 1000 gal' , 'Total Use to This Block of All Bills Passing Through Block, 1000 gal','Cumulative Billed Usage, 1000 gal','Cumulative Billed Usage,%' ), 
          format = 'pipe', linesep = '',
          booktabs = TRUE,
          align = 'c'
  )
```




Plotting the semilogarithmic graph of cumulative billed usage

```{r cumulative billed usage-semilog}
#plotting the semilogarithmic graph of cumulative billed usage

ggplot(data = cumul_bill, aes(x = water_usage)) +
  geom_line(aes(y = cumulative_billed_usage_percent),
            color = "blue",
            linetype = "solid",
            linewidth = 1) +
   scale_x_continuous(
    breaks = c(seq(2, 10, by = 2),
               seq(10, 20, by = 5), 
               seq(20, max(cumul_bill$water_usage), by = 10)),
    trans = "log2"  # Use log base 2 scale for x-axis
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    breaks = seq(0, 100, by = 10)
  ) +
  labs(
    y = "Cumulative Billed Usage Percent",
    x = "Water Usage",
    title = "Cumulative Billed Usage -semilogarithmic"
  ) +
  theme_minimal()
  
  
  
  
```


```{r model1, results='hide'}

#fitting a logistic curve model to the cumulative billed usage
model1 <- nls2( cumulative_billed_usage_percent ~   (a / (1 + b * exp(-c * water_usage))), 
                          data = cumul_bill,
                          start = list(a =1, 
                          b = 5.0, 
                          c = 0.5),
                          algorithm="port",
                        lower = c(a = 0, b = 0, c = 0),
                        upper = c(a = Inf, b = Inf, c = Inf)
                          )

summary(model1)



```



```{r binding model1}
#creating a new column named model1 
cumul_bill<-cumul_bill %>% 
              mutate(model_1=predict(model1, 
                                     newdata = cumul_bill)) %>% 
              mutate(model_1=round(model_1)) %>% 
              mutate(model_1 = structure(model_1, 
                                  label = "Model 1: Logistic Curve"))
```






Plotting original population data and fitted model in a semilogarithmic graph



```{r s curve actual vs prediction}
cumul_bill$predicted <- predict(model1, newdata = cumul_bill)

### Plotting original population data and model 1 in a semilogarithmic graph

ggplot(data = cumul_bill, aes(x = water_usage))+
                geom_line(aes(y = cumulative_billed_usage_percent, color="Actual"), 
                              linetype = "solid", 
                              linewidth = 1)+
  geom_line(data = cumul_bill, aes(y = predicted, color = "Predicted")) + 
  scale_x_continuous(
    breaks = c(seq(2, 10, by = 2),
               seq(10, 20, by = 5), 
               seq(20, max(cumul_bill$water_usage), by = 10)),
    trans = "log2"  # Using log base 2 scale for x-axis
  )+
  labs(x = "Water Usage", y = "Cumulative Billed Usage Percent")+
  scale_y_continuous(breaks = seq(0, 100, by = 5)) +
  theme_minimal()
       

```



Plotting original population data and fitted model in a graph



```{r model vs predicted graph}
cumul_bill$predicted <- predict(model1, newdata = cumul_bill)

### Plotting original population data and model 1 in a graph

ggplot(data = cumul_bill, aes(x = water_usage)) +
                geom_line(aes(y = cumulative_billed_usage_percent, color="Actual"), 
                              linetype = "solid", 
                              linewidth = 1)+
          geom_line(data = cumul_bill, aes(y = predicted, color = "Predicted")) +  
          labs(x = "Water Usage", y = "Cumulative Billed Usage Percent")+
          scale_x_continuous(breaks = seq(0, 80 , by=10 ))+ 
          scale_y_continuous(breaks = seq(0, 100, by = 5)) +
          theme_minimal()
       

```

Calculating Predicted Cumulative Billed Usage Percentage


```{r calculating predicted cum billusage percent}

#Calculating Predicted Cumulative Billed Usage Percentage

predict_cumulative_percent <- function(model, water_usage) {
    
      predicted_percent <- predict(model1, newdata = data.frame(water_usage = water_usage))
  
      return(predicted_percent)
}

water_usage<-10
predicted_percentage <- predict_cumulative_percent(model1, water_usage )
cat("Predicted cumulative billed usage percentage for water usage", water_usage," is:", predicted_percentage, "\n")
```



```{r equation}
# Getting the coefficients from the fitted model (Model 1)
coefficients <- coef(model1)

# Extracting the values of 'a', 'b', and 'c' from the coefficients
a <- coefficients[["a"]]
b <- coefficients[["b"]]
c <- coefficients[["c"]]

# Constructing the equation of the fitted curve
equation <- paste("cumulative billed usage percent =", round(a, 2), "/", "(1 +", round(b, 2), "* exp(-", round(c, 2), "* water usage))")

# The final equation is given as:
cat("Equation of the fitted curve is:", "\n" , equation, "\n")

```
























